---
title: "CTA-ED-exercise3"
author: "Yiru Fang"
date: "2026-03-03"
output: html_document
---

## Setup 

load the packages

```{r, echo=F}
library(kableExtra)
```

```{r, message=F}
library(readr) # more informative and easy way to import data
library(quanteda) # includes functions to implement Lexicoder
library(quanteda.textstats) # for estimating similarity and complexity measures
library(stringdist) # for basic character-based distance measures
library(dplyr) #for wrangling data
library(tibble) #for wrangling data
library(ggplot2) #for visualization
library(tidyr) # for reshaping data (pivot_longer)
```

For this example we'll be using data from the 2017-2018 Theresa May Cabinet in the UK. The data are tweets by members of this cabinet. 

```{r}
tweets  <- readRDS(gzcon(url("https://github.com/cjbarrie/CTA-ED/blob/main/data/comparison-complexity/cabinet_tweets.rds?raw=true")))
```

And we see that the data contain three variables: "username," which is the username of the MP in question; "tweet," which is the text of the given tweet, and "date" in days in yyyy-mm-dd format. 

```{r}
head(tweets)
```

And there are 24 MPs whose tweets we're examining. 

```{r}
unique(tweets$username)

length(unique(tweets$username))
```

## Generate document feature matrix

In order to use the `quanteda` package and its accompanying `quanteda.textstats` package, we need to reformat the data into a quanteda "corpus" object. To do this we just need to specify the text we're interested in as well as any associated document-level variables in which we're interested. 

We can do this as follows. 

```{r}
#make corpus object, specifying tweet as text field
tweets_corpus <- corpus(tweets, text_field = "tweet")

#add in username document-level information
docvars(tweets_corpus, "username") <- tweets$username

tweets_corpus
```

reformat the data into a document feature matrix.

```{r}
dfmat <- dfm(tokens(tweets_corpus, remove_punct = TRUE)) %>%
  dfm_remove(pattern = stopwords("english"))

dfmat
```


## Exercise 1
Compute Euclidean and Manhattan distances between MPs and Theresa May

```{r}
# Group tweets by username so that each MP becomes one document
dfmat_grouped <- dfm_group(dfmat,
                           groups = docvars(dfmat, "username"))

# Convert counts to proportions
dfmat_grouped <- dfm_weight(dfmat_grouped, scheme = "prop")
```
We aggregate all tweets by each MP into a single document so that we compare politicians rather than individual tweets.

```{r}
# Compute Euclidean distance between documents
euclid_dist <- textstat_dist(dfmat_grouped,
                             margin = "documents",
                             method = "euclidean")
```
Euclidean distance measures straight-line distance between document vectors in high-dimensional space.
Higher values = more different language use.

```{r}
# Compute Manhattan distance between documents
manhattan_dist <- textstat_dist(dfmat_grouped,
                                margin = "documents",
                                method = "manhattan")
```
Manhattan distance sums the absolute differences between word frequencies across documents.
It is often more robust to outliers than Euclidean distance.

```{r}
## Extract Theresa May comparisons
# Convert to matrix
euclid_mat <- as.matrix(euclid_dist)
manhattan_mat <- as.matrix(manhattan_dist)

# Find row corresponding to Theresa May
may_row <- which(rownames(euclid_mat) == "theresa_may")

# Extract distances between Theresa May and other MPs
euclid_may <- euclid_mat[may_row, -may_row]

# Find row corresponding to Theresa May
may_row <- which(rownames(manhattan_mat) == "theresa_may") 

# Extract distances between Theresa May and other MPs
manhattan_may <- manhattan_mat[may_row, -may_row]

# Create dataframe
other_users <- rownames(euclid_mat)[-may_row]

dist_df <- data.frame(
  username = other_users,
  euclidean = euclid_may,
  manhattan = manhattan_may
)
```

```{r}
# Identify most similar MP
dist_df %>% arrange(euclidean) %>% head()

# Identify most different MP
dist_df %>% arrange(desc(euclidean)) %>% head()
```

```{r}
dist_long <- dist_df %>%
  pivot_longer(cols = c(euclidean, manhattan),
               names_to = "method",
               values_to = "distance")

ggplot(dist_long, aes(x=reorder(username, distance), y=distance, color=method)) +
  geom_point() +
  coord_flip()
```


## Exercise 2
Estimate at least three additional complexity measures

The article by Schoonvelde et al. (2019) uses Flesch-Kincaid. We now compare other readability measures.

```{r}
# Assuming have a character vector of texts:
texts <- as.character(tweets_corpus)
```

```{r}
# Tokenise and remove punctuation marks (keep only the words)
toks_all <- tokens(texts, remove_punct = TRUE)
```

```{r}
# a) Type-Token Ratio Calculate the TTR for each document as follows: TTR = Number of unique words / Total number of words
ttr_scores <- sapply(toks_all, function(toks) {
  tokens_n <- length(toks)                # Total Word Count
  if (tokens_n == 0) return(NA)           # Avoid division by zero in empty documents
  types <- length(unique(toks))           # Unique Word Count
  types / tokens_n                        # TTR
})
```

```{r}
# b) Shannon Entropy
shannon_entropy <- function(toks) {
  tokens_n <- length(toks)
  if (tokens_n == 0) return(NA)
  
  freqs <- table(toks) / tokens_n        # Frequency Probability
  -sum(freqs * log(freqs))               # Entropy Formula
}

# Applied to all documents
entropy_scores <- sapply(toks_all, shannon_entropy)
```

```{r}
# c) Average Word Length
avg_word_length <- function(toks) {
  tokens_n <- length(toks)
  if (tokens_n == 0) return(NA)
  
  mean(nchar(toks))                      # Average Character Count
}

avg_len_scores <- sapply(toks_all, avg_word_length)
```

```{r}
# Create a corpus object (containing document-level variables)
corp <- corpus(texts)

# Calculate the Flesch-Kincaid readability score
# measure (optional) "Flesch.Kincaid" or "Flesch"
fk_scores <- textstat_readability(corp, measure = "Flesch.Kincaid")

# Extract the score vector
flesch_kincaid <- fk_scores$Flesch.Kincaid
```

```{r}
# summary 
complexity_df <- data.frame(
  TTR = ttr_scores,
  Shannon_Entropy = entropy_scores,
  Avg_Word_Length = avg_len_scores,
  Flesch_Kincaid = flesch_kincaid   # Newly Added Comparative Indicators
)

# Check the previous few lines
head(complexity_df)

summary(complexity_df)
```


```{r}
# Correlation analysis: Calculate the correlation coefficient between other indicators and Flesch-Kincaid.
cor_matrix <- cor(complexity_df, use = "complete.obs")
print(cor_matrix)

# Visualization: Creating a scatter plot matrix
pairs(complexity_df, pch = 19, col = rgb(0,0,1,0.5))
```


```{r}
# Compare MPs vs Theresa May
is_pm <- grepl("Theresa_May", docnames(tweets_corpus))

aggregate(complexity_df, 
          by = list(Group = ifelse(is_pm, "PM", "MP")), 
          mean, na.rm = TRUE)
```


3. (Optional) Similarity Scores by Week
```{r}
# Step A — Add week variable
library(lubridate)                # Date Processing

# extract metadata
meta <- docvars(tweets_corpus)

# create week variable
meta$week <- floor_date(as.Date(meta$date), unit = "week")

# assign back to corpus
docvars(tweets_corpus) <- meta
```

```{r}
# Step B — Create Clean dfm
# Tokenize and remove punctuation, symbols and numbers.
toks <- tokens(tweets_corpus,
               remove_punct = TRUE,
               remove_symbols = TRUE,
               remove_numbers = TRUE)

# Create DFM (Automatically Calculate Word Frequency)
dfm_all <- dfm(toks)

# remove rare words
dfm_all <- dfm_trim(dfm_all, min_termfreq = 5)

head(docvars(dfm_all))
```

```{r}
# Step C — Loop Through Weeks & Compute Cosine Similarity
library(proxy)

# Obtain all unique weeks (sorted)
weeks <- sort(unique(docvars(dfm_all)$week))

# Initialize the result data frame
results <- data.frame(
  week = weeks,
  cosine_similarity = NA_real_   # placeholder
)

# For each weekly cycle
for (i in seq_along(weeks)) {
  
  current_week <- weeks[i]
  
  # subset to this week
  dfm_week <- dfm_subset(dfm_all, week == current_week)
  
  # split PM vs MPs
  # Use dfm_group to combine the documents within the same group into a "virtual document"
  dfm_grouped <- dfm_group(dfm_week,
                            groups = ifelse(docvars(dfm_week)$username == "theresa_may",
                                            "PM", "MP"))
  
  # Only when both groups are present (that is, there are both PM posts and MPs posts in this week) will the similarity be calculated.
  if (ndoc(dfm_grouped) == 2) {
    # Calculate the cosine similarity between the two documents (PM vector and MP vector)
    sim <- textstat_simil(dfm_grouped, method = "cosine")
    # Extract the similarity value (the element in the first row and second column of the matrix)
    results$cosine_similarity[i] <- as.numeric(sim[1,2])
  }
  # Otherwise, keep it as NA (indicating that the calculation cannot be performed for this week)
}
```

```{r}
ggplot(results, aes(x = week, y = cosine_similarity)) +
  geom_line(color = "steelblue", size = 1) +          # Line Chart
  geom_point(color = "steelblue", size = 2) +         # dot mark
  labs(title = "Cosine similarity between the tweets of MPs and Theresa May on a weekly basis",
       x = "weeks", y = "cosine similarity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate the x-axis labels to avoid overlapping
```

```{r}
summary(results$cosine_similarity)
sum(!is.na(results$cosine_similarity))
```

